---
title: "Spacial Data"
output: html_document
---
```{r, echo = FALSE}
library(knitr)
opts_chunk$set(warning=FALSE, message=FALSE, fig.align='center') 
```

# Introduction



**TAKE AWAY POINTS FROM THIS POST**

* 

* 

* 


# Data

First we read in and organize the data.

```{r}
library(dplyr)
library(ggmap)

# data = read.table("data_locations.txt", header=T, sep="\t") %>%
#   mutate(address = as.character(address)) %>%
#   rowwise() %>%
#   mutate(longitude = geocode(address, source = "google")[1, 1]) %>%
#   mutate(latitude = geocode(address, source = "google")[1, 2])

data = read.table("data_saved.txt", header=T, sep="\t")
```

Let's view the data.

```{r, echo = FALSE}
library(DT)

datatable(data, options = list(pageLength = 12, dom = 'tip'), rownames = FALSE)
```

We then create a map object pulled from Google maps. 

```{r, cache=TRUE}
# Get city specific maps
paris_map = get_googlemap(center = "Paris", maptype = "roadmap", zoom = 12, size = c(640, 420))

berlin_map = get_googlemap(center = "Berlin", maptype = "roadmap", zoom = 11, size = c(640, 420))

barcelona_map = get_googlemap(center = "Barcelona", maptype = "roadmap", zoom = 12, size = c(640, 420))

prague_map = get_googlemap(center = "Prague", maptype = "roadmap", zoom = 12, size = c(640, 420))
```

With our map object saved 

```{r}
# Plot city specific maps with landmarks noted
city_plot = function(city_name, city_map){
  ggmap(city_map, extent = "device") +
  geom_point(data = subset(data, city == city_name), aes(x = longitude, y = latitude),
             color = "red", size = 4)
}

paris.plot = city_plot("Paris", paris_map)
paris.plot

berlin.plot = city_plot("Berlin", berlin_map)
berlin.plot

barcelona.plot = city_plot("Barcelona", barcelona_map)
barcelona.plot

prague.plot = city_plot("Prague", prague_map)
prague.plot
```


# Delany Tessilation Area

With our maps and data points in place let's compute the tessilation for each map.

```{r}
library(deldir)

# data_deldir = data %>%
#   group_by(city) %>%
#   do(deldir(.$longitude, .$latitude)) %>%
#   ungroup()
# 
# 
# data_deldir = data %>%
#   nest(-city) %>%
#   mutate(col = map2(longitude, latitude, deldir)) %>%
#   select(city, col) %>%
#   unnest()

### TO BE CHANGED
paris_data = filter(data, city == "Paris")
paris_deldir = deldir(paris_data$longitude, paris_data$latitude)

berlin_data = filter(data, city == "Berlin")
berlin_deldir = deldir(berlin_data$longitude, berlin_data$latitude)

barcelona_data = filter(data, city == "Barcelona")
barcelona_deldir = deldir(barcelona_data$longitude, barcelona_data$latitude)

prague_data = filter(data, city == "Prague")
prague_deldir = deldir(prague_data$longitude, prague_data$latitude)
```

Now we can update our figures with the tessliations.

```{r}
del_plot = function(city_data, city_map) {
  city_map +
  geom_segment(data = city_data$delsgs, aes(x = x1, y = y1, xend = x2, yend = y2),
    size = 1, color= "black")
}

paris_del.plot = del_plot(paris_deldir, paris.plot)
paris_del.plot

berlin_del.plot = del_plot(berlin_deldir, berlin.plot)
berlin_del.plot

barcelona_del.plot = del_plot(barcelona_deldir, barcelona.plot)
barcelona_del.plot

prague_del.plot = del_plot(prague_deldir, prague.plot)
prague_del.plot
```

Let's compare the areas. The area for Paris is `r paris_deldir$del.area`, Berlin `r berlin_deldir$del.area`, Barelona `r barcelona_deldir$del.area`, and Prague `r prague_deldir$del.area`.


# Delany Tessilation Ceintroid

Caculate centroid of the tesselation.

```{r}
## TO BE CHANGED
paris_deldir_sum = data.frame(paris_deldir$summary) %>%
  mutate(city = "Paris")

berlin_deldir_sum = data.frame(berlin_deldir$summary) %>%
  mutate(city = "Berlin")

barcelona_deldir_sum = data.frame(barcelona_deldir$summary) %>%
  mutate(city = "Barcelona")

prague_deldir_sum = data.frame(prague_deldir$summary) %>%
  mutate(city = "Prague")

data_deldir_sum = paris_deldir_sum %>%
  bind_rows(berlin_deldir_sum) %>%
  bind_rows(barcelona_deldir_sum) %>%
  bind_rows(prague_deldir_sum)

data_deldir_centroid = data_deldir_sum %>%
  group_by(city) %>%
  summarise(cent_x = sum(x * del.wts),
            cent_y = sum(y * del.wts)) %>%
  ungroup()
```

Plot figures with centroid noted.

```{r}
# Make plots with centroid
cent_plot = function(city_name, city_map){
  city_map +
  geom_point(data = subset(data_deldir_centroid, city == city_name),
              aes(x = cent_x, y = cent_y),
              size = 8, color= "blue")
    
}

paris_del_cent.plot = cent_plot("Paris", paris_del.plot)
paris_del_cent.plot

berlin_del_cent.plot = cent_plot("Berlin", berlin_del.plot)
berlin_del_cent.plot

barcelona_del_cent.plot = cent_plot("Barcelona", barcelona_del.plot)
barcelona_del_cent.plot

prauge_del_cent.plot = cent_plot("Prague", prague_del.plot)
prauge_del_cent.plot
```


# Conclusion





